---
title: "CDP Technical Task - Junior Data Analyst"
author: "Isabella Morgante"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/CDP_Task/Code/") # set working directory
rm(list=ls(all=TRUE))
graphics.off()
require(ggplot2)
#require(gridExtra)
```

This report documents the approach used to explore the electricity generation by source for the top 5 power generating countries over time. 
The dataset was first inspected and assessed for quality and then each analysis question was addressed. 

### Inspecting the Data 

Load the data using the appropriate file path.

```{r}
cdp_data <- read.csv(file = "../Data/Electricity generation by source - top 5.csv", stringsAsFactors = TRUE)
```
Examine content and structure of the dataset.
```{r}
head(cdp_data) # View first few rows
str(cdp_data) # View data structure

# Change column name to avoid dots
names(cdp_data)[names(cdp_data) == "Electricity.Generation..GWh."] <- "ElectricityGeneration_GWh"
```
The output shows the data contains information about the electricity generation by source and country from 1990-2019. There are 6 *Country* categories representing the top 5 power generating countries and global values. There are 10 types of electricity generation listed under the *Sources* column. The electricity generation units are consistent throughout the data as the *Units* column has only one factor, GWh. 

The electricity generation data contains missing values (NAs). For the purpose of this exercise, the assumption has been made that the  missing values are due to a country not generating electricity by a particular source for that year. Following this assumption, all rows containing missing electricity generation data have been omitted from analysis.

```{r}
names(which(colSums(is.na(cdp_data))>0)) # List columns containing NAs
head(cdp_data[rowSums(is.na(cdp_data)) > 0, ]) # List first rows that contain NAs
###### FIND WAY TO REPESENT THIS AS GRAPH NOT JUST AS A TABLE.  REPORT NUMBER OF NA rows
```

### Question 1: Total Electricity Generation

*Calculate the total electricity generation per year. How does this change overtime?*

To find the total electricity generated by each country per year, the amount of generation from each source is summed.

```{r}
total_electricity <- aggregate(cdp_data$ElectricityGeneration_GWh, by=list(Year = cdp_data$Year, Country = cdp_data$Country),FUN=sum, na.rm=TRUE) # Sum sources
names(total_electricity)[names(total_electricity) == "x"] <- "ElectricityGeneration_GWh" #Name column
```
The total electricity generation by top 5 power generating countries over 30 years show some interesting tends (see below figures). Japan and the Russian Federation show little change in the total amount of electricity generation and the United States' electricity generation appears to have plateaued since 2005. Both India and China have seen an increase in electricity generation.

From the figure, China showed the most dramatic change in their electricity generation, with an increasing rate of generation. I have selected China as my country of choice for further investigation.

**DO FURTHER ANALYSIS HERE -- % change since 1990? see what type of growth?  HOW DOES THIS CHANGE OVER TIME**

```{r, echo=FALSE}
country_total_electricity <- total_electricity[!total_electricity$Country=='World',] # ignore world values

#Plots for countries
p_country_totals <- ggplot(country_total_electricity, aes(x = Year, y = ElectricityGeneration_GWh, fill = Country)) + 
  geom_point(aes(color=Country)) +
  geom_line(aes(color=Country))+ 
  theme_bw() + 
  #theme(legend.position="bottom") +
  ylab("Total Electricity Generation [GWh]") 
  
p_country_totals
```

### Question 2: Fossil Fuel Generation
*What percentage of electricity is generated from fossil fuel sources per year?*

First the original data is subset by the country of interest (in this analysis China has been selected).
```{r}
# Choose country to explore
country <- "China"
country_data <- subset(cdp_data, cdp_data$Country == country) # Subset data by country of interest
country_data <- na.omit(country_data) # remove NAs 
```


The amount of electricity generation by different sources in China over time can be seen in the figure below. The majority of electricity is generated by coal in China followed by natural gas, the other sources of electricity represent a small proportion of overall generation.

```{r, echo=FALSE}
p_sources <- ggplot(country_data, aes(x = Year, y = ElectricityGeneration_GWh, fill = Source)) + 
  geom_point(aes(color=Source)) +
  geom_line(aes(color=Source))+ 
  theme_bw() +
  ylab("Electricity Generation [GWh]") 

p_sources
```

To explore how the amount of electricity generated by fossil fuels have changed over time, a the proportion of electricity generated by fossil fuels (oil, coal and natural gas) with respect to total electricity generation is found.

First the combined amount of electricity generated by fossil fuels per year is found. 

```{r}
FF_electricity <- subset(country_data, country_data$Source == "Coal" | country_data$Source == "Natural gas" | country_data$Source == "Oil") # Subset for fossil fuels

FF_electricity <- aggregate(FF_electricity$ElectricityGeneration_GWh, by=list(Year = FF_electricity$Year), FUN=sum, na.rm=TRUE)  # Sum electricity generation

names(FF_electricity)[names(FF_electricity) == "x"] <- "ElectricityGeneration_GWh"
FF_electricity$Source <- "Fossil Fuel"
```

Then the amount of fossil fuel electricity generation is found as a percentage of total generation and is recorded in a new dataframe.

```{r}
country_total <- subset(total_electricity, total_electricity$Country == country) # Subset total emissions data by selected country
country_total$Country <- NULL # Remove county column
country_total$Source <- "Total" # Add Source 

FF_totals <- rbind(country_total, FF_electricity) # Combine fossil fuel and total values
percent_FF <- data.frame(matrix(ncol=2,nrow=length(unique(FF_totals$Year)), dimnames=list(NULL, c("Year", "Percentage")))) # Create empty dataframe to store percentage calculations 
count <- 1

# Calculate percentages per year 
for (i in unique(FF_totals$Year)){
  df <- subset(FF_totals, FF_totals$Year == i)
  percent <- df[df$Source == "Fossil Fuel", "ElectricityGeneration_GWh"] / df[df$Source == "Total", "ElectricityGeneration_GWh"]
  
  percent_FF$Year[count] <- i
  percent_FF$Percentage[count] <- percent
  
  count = count + 1
  }
```

The percentage of electricity generated by fossil fuels in China can be seen in the figure below.  From about 2000, the proportion of China's electricity that is generated by fossil fuels has been steadily declining.  However, despite these reductions, China is still producing the majority of its electricity by fossil fuels.

```{r, echo=FALSE}
p_FF_percent <- ggplot(percent_FF, aes(x = Year, y = Percentage)) + 
  geom_point() +
  geom_line()+ 
  theme_bw() +
  ylab("Proportion of electricity generated by fossil fuels") 

p_FF_percent
```

### Question 3: Coal Generation
*For the most recent year, what percentage of world coal generation do the top five countries represent? How does this compare to natural gas?*



### Question 4: Total GHG Lifecycle Emissions
*Using data from the previous exercise with the table you have just extracted, calculate the total GHG lifecycle emissions per year for a country of your choice using the provided equation.*

The ‘Lifecycle greenhouse gas emissions by electricity source’ table was extracted from <https://en.wikipedia.org/wiki/Emission_intensity> using `rvest` package and the appropriate XPath was found by inspecting the source code of the Wikipedia webpage.

```{r}
require(rvest)

url <- "https://en.wikipedia.org/wiki/Emission_intensity"
table_XPath <- '//*[@id="mw-content-text"]/div[1]/table[1]'

lifecycle_GHG_emissions <- url %>%
  read_html() %>%
  html_nodes(xpath=table_XPath) %>%
  html_table()
lifecycle_GHG_emissions <-lifecycle_GHG_emissions[[1]]
```
```{r}
lifecycle_GHG_emissions
```

